A Software Engineering Perspective on \\ OpenMP and PThreads

### 2017-08-09 sections/01-introduction.tex
More text Revisions.

The contribution of this paper is a comparison between \textit{Pthreads} and
\textit{OpenMP}, two commonly used APIs for parallel and distributed
programming.  The comparison considers the perceptions and performance results
from undergraduate and graduate students enrolled in the course of Concurrent,
Parallel and Distributed Computing of the University of São Paulo, Brazil.  We
show that students were able to achieve better performance results using
\textit{Pthreads} than \textit{OpenMP}, despite their perception that
\textit{Pthreads} is harder to learn and use.

The paper is organized as follows.
Section \ref{sec:apis} describes \textit{OpenMP} and \textit{Pthreads}.
Section \ref{sec:background} discusses related work.
Section \ref{sec:researchdesign} discusses the course of
Concurrent, Parallel and Distribute Computing of the University
of São Paulo, and presents the questionnaire and our research
questions.
Section \ref{sec:responses} presents, discusses and analyses
the students' responses to the questionnaire.
Section \ref{sec:conclusion} summarizes the results and
presents future work.

>>>>

The contribution of this paper is a comparison between \textit{Pthreads} and
\textit{OpenMP}, two commonly used APIs for parallel and distributed
programming.  The comparison considers the perceptions and performance results
from undergraduate and graduate students enrolled in the course of Concurrent,
Parallel and Distributed Computing of the University of São Paulo, Brazil.  We
show that students were able to achieve better performance results with
\textit{Pthreads} than with \textit{OpenMP}, despite their perception
that \textit{Pthreads} is harder to learn and use.

The paper is organized as follows.
Section \ref{sec:apis} briefly describes \textit{OpenMP} and \textit{Pthreads}.
Section \ref{sec:background} discusses related work.
Section \ref{sec:researchdesign} presents the course of
Concurrent, Parallel and Distribute Computing of the University
of São Paulo, and introduces the questionnaire and our research
questions.
Section \ref{sec:responses} presents, discusses and analyses
the students' responses to the questionnaire.
Section \ref{sec:conclusion} summarizes the results and
presents future work.

### 2017-08-09 sections/02-apis.tex
More text Revisions.

The C/C++ \textit{Pthreads} library controls thread execution in GNU/Linux by
creating and synchronizing threads. In this model, the execution begins in a
sequential region and threads must be explicitly started to begin a parallel
region.  Listing~\ref{lst:pthreads} presents a \textit{Pthreads}
example. The code shows the parallel computation of the dot product of two
arrays.

@@

>>>>

The \textit{Pthreads} library controls thread execution by
creating and synchronizing threads. In this model, the execution begins in a
sequential region and threads must be explicitly started to begin a parallel
region.  Listing~\ref{lst:pthreads} presents a \textit{Pthreads}
example. The code shows the parallel computation of the dot product of two
arrays.

@@

Note that the examples shown in Listings \ref{lst:pthreads} and
\ref{lst:omp} solve the same problem. The example in \textit{Phthreads}
contains much more code than the example in \textit{OpenMP}, and seems
to be much more complex as well. It is therefore understandable that students might
feel overwhelmed by \textit{Pthreads} example, especially when both examples are
presented side-by-side.

### 2017-08-09 sections/04-research-design.tex
More text Revisions.

This section describes our research design. We devised a questionnaire
regarding student experience with solving the same problem using
\textit{Pthreads} and \textit{OpenMP} during the Concurrent, Parallel and
Distributed Computing Course at the Computer Science program of the University
of São Paulo. We answered 3 research questions using analyses of the students'
responses.

\subsection{The Concurrent, Parallel and Distributed Computing Course}

@@

The course presents students with a series of software engineering challenges
in the form of assignments. The final is composed of an essay and a
presentation on a relevant topic related to Concurrent, Parallel and
Distributed Computing.

We devised an assignment to measure the students' difficulties to
learn and use \textit{Pthreads} and \textit{OpenMP} to solve
a simple problem. Students had to implement parallel versions
of a sequential code using \textit{Pthreads} and \textit{OpenMP}.
We provided a sequential implementation for the calculation of
the Mandelbrot Set. The sequential code presented an embarrassingly
parallel problem with two nested loops, and a third loop with
dependent iterations that could not be parallelized easily.

@@

We obtained responses, and consent to use them in our research, from 38 of the
54 students enrolled in the Concurrent, Parallel and Distributed Computing
course. Section \ref{sec:background} presents a detailed description of
each set of questions our analysis of the students' responses.

\subsection{Research Questions}

In previous work~\cite{goncalves:OpenMPNotEasy} we showed that
teaching \textit{OpenMP} requires teaching important parallel
and distributed computing concepts first, since it is easy
to find errors on commonly used \textit{OpenMP} learning material.

>>>>

This section describes our research design. We devised a questionnaire
regarding student experience with solving the same problem using
\textit{Pthreads} and \textit{OpenMP} during the Concurrent, Parallel and
Distributed Computing Course at the Computer Science program of the University
of São Paulo. We answered 3 research questions analysing the students'
responses.

\subsection{The Course}

@@

The course presents students with a series of software engineering challenges
in the form of assignments. The final is composed of an essay and a
presentation on a relevant topic related to Concurrent, Parallel and
Distributed Computing.  The course material is available
online~\footnote{\url{https://phrb.github.io/MAC5742-0219} [Accessed in
07/08/2017]}.

We devised an assignment to measure the students' difficulties to learn and use
\textit{Pthreads} and \textit{OpenMP} to solve a simple problem. Students had
to implement parallel versions of a sequential code using \textit{Pthreads} and
\textit{OpenMP}.  We provided a sequential implementation for the calculation
of the Mandelbrot Set~\cite{douady1984etude}. The sequential code presented an
embarrassingly parallel problem with two nested loops, and a third loop with
dependent iterations that could not be parallelized easily.

@@

We obtained responses, and consent to use them in our research, from 38 of the
54 students enrolled in the Concurrent, Parallel and Distributed Computing
course. Section \ref{sec:responses} presents a detailed description of
each set of questions and our analysis of the students' responses.

\subsection{Research Questions}
\label{sec:resques}

In previous work~\cite{goncalves:OpenMPNotEasy} we showed that teaching
\textit{OpenMP} requires the introduction of important parallel and distributed
computing concepts first, since it is easy to find errors on commonly used
\textit{OpenMP} learning material.

### 2017-08-09 sections/05-responses.tex
Change title and minor edits.
Minor revisions and spacing changes.
More text Revisions.

\todo[inline,color=cyan,author=Pedro]{Improve figure colorscheme}

@@

The course of Concurrent, Parallel and Distributed Computing is offered to
students of both the graduate and undergraduate programs in Computer Science at
the University of São Paulo. To gauge the proficiency and knowledge of our
students we elaborated a five-point scale from \textit{Do not Know} $(1)$ to
\textit{Expert or Proficient} $(5)$.  We asked the students to express their
proficiency level on this scale, and their responses are shown in Figure
\ref{fig:background}.

@@

We asked the students to evaluate their proficiency in \textit{OpenMP} and
\textit{Pthreads}, \textit{before} completing the assignment, with two sets of
questions. The first set used the same five-point scale as the other background
questions and is also shown in Figure \ref{fig:background}. The students
reported significantly larger proficiency in \textit{Pthreads} than
\textit{OpenMP}, and the majority of the students reported they did not know
\textit{OpenMP} before the course.

\begin{table}[htpb]
    \centering
    \begin{tabular}{@{}p{0.9\columnwidth}p{0.08\columnwidth}@{}}
        \toprule
        \multicolumn{1}{c}{\scriptsize{Have you$\dots$}} & \textnumero \\ \midrule
        \scriptsize{Had contact with parallel and distributed concepts before?} & $(1)$ \\
        \scriptsize{Had contat with the APIs before?} & $(2)$ \\
        \scriptsize{Learned from the OpenMP class?} & $(3)$ \\
        \scriptsize{Learned from the Pthreads class?} & $(4)$ \\ \bottomrule
    \end{tabular}
    \caption{Questions for Figure \ref{fig:classes}}
    \label{tab:classes}
\end{table}

\begin{figure}[htpb]
    \centering
    \includegraphics[width=0.85\columnwidth]{classes_questions}
    \caption{Students' responses to questions in Table \ref{tab:classes}}
    \label{fig:classes}
\end{figure}

@@

\begin{table}[htpb]
    \centering
    \begin{tabular}{@{}p{0.7\columnwidth}p{0.1\columnwidth}p{0.05\columnwidth}@{}}
        \toprule
        \multicolumn{2}{c}{\scriptsize{It is easy to$\dots$}} & \textnumero \\ \midrule
        \multirow{2}{*}{\parbox{0.7\columnwidth}{\scriptsize{Parallelize loops with independent iterations using:}}} & \scriptsize{OpenMP} & $(1)$ \\
        & \scriptsize{Pthreads} & $(2)$ \\
        \addlinespace{}
        \multirow{2}{*}{\parbox{0.7\columnwidth}{\scriptsize{Parallelize nested loops with independent iterations using:}}} & \scriptsize{OpenMP} & $(3)$ \\
        &  \scriptsize{Pthreads} & $(4)$ \\
        \addlinespace{}
        \multirow{2}{*}{\parbox{0.7\columnwidth}{\scriptsize{Improve the performance of sequential code using:}}} & \scriptsize{OpenMP} & $(5)$  \\
        &  \scriptsize{Pthreads} & $(6)$ \\ \bottomrule
    \end{tabular}
    \caption{Likert-like Scale questions for Figure \ref{fig:likert}}
    \label{tab:likert}
\end{table}

\begin{figure}[htpb]
    \centering
    \includegraphics[width=0.85\columnwidth]{likert_questions}
    \caption{Students' responses to the Likert Scale questions in Table \ref{tab:likert}}
    \label{fig:likert}
\end{figure}

@@

\begin{table}[htpb]
    \centering
    \begin{tabular}{@{}p{0.9\columnwidth}p{0.08\columnwidth}@{}}
        \toprule
        \multicolumn{1}{c}{\scriptsize{Which technology$\dots$}} & \textnumero \\ \midrule
        \scriptsize{Was more difficult to learn?} & $(7)$ \\
        \scriptsize{Was simpler to use?} & $(8)$ \\
        \scriptsize{Was easier to use?} & $(9)$ \\
        \scriptsize{Was easier to use for parallelizing loops with independent iterations?} & $(10)$ \\
        \scriptsize{Was easier to use for parallelizing nested loops with independent iterations?} & $(11)$  \\
        \scriptsize{Was easier to use for improving the performance of a sequential program?} & $(12)$  \\
        \scriptsize{Had the best performance on your assignment?} & $(13)$ \\ \bottomrule
    \end{tabular}
    \caption{Questions for Figure \ref{fig:comparisons}}
    \label{tab:comparisons}
\end{table}

\begin{figure}[htpb]
    \centering
    \includegraphics[width=0.85\columnwidth]{comparisons}
    \caption{Students' responses to the questions in Table \ref{tab:comparisons}}
    \label{fig:comparisons}
\end{figure}

>>>>

%\todo[inline,color=cyan,author=Pedro]{Improve figure colorscheme}

@@

To gauge the proficiency and knowledge of our students we elaborated a
five-point scale from \textit{Do not Know} $(1)$ to \textit{Expert or
Proficient} $(5)$.  We asked the students to express their proficiency level on
this scale, and their responses are shown in Figure \ref{fig:background}.

@@

We asked the students to evaluate their proficiency in \textit{OpenMP} and
\textit{Pthreads}, \textit{before} completing the assignment, with two sets of
questions. The first set used the same five-point scale as the other background
questions and is also shown in Figure \ref{fig:background}. The students
reported significantly larger proficiency in \textit{Pthreads} than in
\textit{OpenMP}, and the majority of the students reported they did not know
\textit{OpenMP} before the course.

\begin{table}[htpb]
    \centering
    \begin{tabular}{@{}p{0.9\columnwidth}p{0.08\columnwidth}@{}}
        \toprule
        \multicolumn{1}{c}{\scriptsize{Have you$\dots$}} & \textnumero \\ \midrule
        \scriptsize{Had contact with parallel and distributed concepts before?} & $(1)$ \\
        \scriptsize{Had contat with the APIs before?} & $(2)$ \\
        \scriptsize{Learned from the OpenMP class?} & $(3)$ \\
        \scriptsize{Learned from the Pthreads class?} & $(4)$ \\ \bottomrule
    \end{tabular}
    \caption{Questions for Figure \ref{fig:classes}}
    \label{tab:classes}
    \vspace{-0.6cm}
\end{table}

\begin{figure}[htpb]
    \vspace{-0.7em}
    \centering
    \includegraphics[width=0.85\columnwidth]{classes_questions}
    \caption{Students' responses to questions in Table \ref{tab:classes}}
    \label{fig:classes}
    \vspace{-1em}
\end{figure}

@@

\begin{table}[htpb]
    \centering
    \begin{tabular}{@{}p{0.7\columnwidth}p{0.1\columnwidth}p{0.05\columnwidth}@{}}
        \toprule
        \multicolumn{2}{c}{\scriptsize{It is easy to$\dots$}} & \textnumero \\ \midrule
        \multirow{2}{*}{\parbox{0.7\columnwidth}{\scriptsize{Parallelize loops with independent iterations using:}}} & \scriptsize{OpenMP} & $(1)$ \\
        & \scriptsize{Pthreads} & $(2)$ \\
        \addlinespace{}
        \multirow{2}{*}{\parbox{0.7\columnwidth}{\scriptsize{Parallelize nested loops with independent iterations using:}}} & \scriptsize{OpenMP} & $(3)$ \\
        &  \scriptsize{Pthreads} & $(4)$ \\
        \addlinespace{}
        \multirow{2}{*}{\parbox{0.7\columnwidth}{\scriptsize{Improve the performance of sequential code using:}}} & \scriptsize{OpenMP} & $(5)$  \\
        &  \scriptsize{Pthreads} & $(6)$ \\ \bottomrule
    \end{tabular}
    \caption{Likert-like Scale questions for Figure \ref{fig:likert}}
    \label{tab:likert}
    \vspace{-0.6em}
\end{table}

\begin{figure}[htpb]
    \vspace{-2em}
    \centering
    \includegraphics[width=0.85\columnwidth]{likert_questions}
    \caption{Students' responses to the Likert Scale questions in Table \ref{tab:likert}}
    \label{fig:likert}
    \vspace{-1em}
\end{figure}

@@

\begin{table}[htpb]
    \centering
    \begin{tabular}{@{}p{0.9\columnwidth}p{0.08\columnwidth}@{}}
        \toprule
        \multicolumn{1}{c}{\scriptsize{Which technology$\dots$}} & \textnumero \\ \midrule
        \scriptsize{Was more difficult to learn?} & $(7)$ \\
        \scriptsize{Was simpler to use?} & $(8)$ \\
        \scriptsize{Was easier to use?} & $(9)$ \\
        \scriptsize{Was easier to use for parallelizing loops with independent iterations?} & $(10)$ \\
        \scriptsize{Was easier to use for parallelizing nested loops with independent iterations?} & $(11)$  \\
        \scriptsize{Was easier to use for improving the performance of a sequential program?} & $(12)$  \\
        \scriptsize{Had the best performance on your assignment?} & $(13)$ \\ \bottomrule
    \end{tabular}
    \caption{Questions for Figure \ref{fig:comparisons}}
    \label{tab:comparisons}
    \vspace{-0.6cm}
\end{table}

\begin{figure}[htpb]
    \vspace{-.7em}
    \centering
    \includegraphics[width=0.85\columnwidth]{comparisons}
    \caption{Students' responses to the questions in Table \ref{tab:comparisons}}
    \label{fig:comparisons}
    \vspace{-1em}
\end{figure}

### 2017-08-09 sections/07-conclusion.tex
More text Revisions.

In this paper we elaborated a questionnaire to gauge students' perceptions and
measure their difficulty to learn \textit{OpenMP} and \textit{Pthreads}.  The
questionnaire contained questions that regarded each technology individually
and in comparison with the other.  The students were enrolled in the
Concurrent, Parallel and Distributed course from the University of São Paulo,
Brazil.

The students' responses to the questionnaire enabled conclude that the students
perceived less difficulty using and learning \textit{OpenMP}, in the context of
a course assignment, but that they also achieved greater performance
improvements using \textit{Pthreads}.

@@

The perceptions of usability and performance improvement regarding
\textit{OpenMP} was not reflected in the actual implementations and
submissions.  We assert that \textit{OpenMP} is still not as easy as it appears
to be, and that the students' responses strengthen the argument for teaching
lower-level technologies for parallel and distributed computing.
The main contribution of this paper is the conclusion that students
are able to achieve better performance results using \textit{Pthreads}
than \textit{OpenMP}, despite their perception that \textit{Pthreads}
is harder to learn and use.

In future work we will compare the perception and performance of students
regarding learning and achieving performance improvements using different
languages and tools for parallel and distributed computing. We will also study
how attempting solving more difficult parallel and distributed problem impacts
the students' perceptions and performance.

>>>>

In this paper we elaborated a questionnaire to gauge students' perceptions and
measure their difficulty to learn \textit{OpenMP} and \textit{Pthreads}.  The
questionnaire contained questions that regarded each technology individually
and in comparison with the other.

The students' responses to the questionnaire enable the conclusion that the
students perceived less difficulty using and learning \textit{OpenMP}, in the
context of a course assignment, but that they also achieved greater performance
improvements using \textit{Pthreads}.

@@

The perceptions of usability and performance improvement regarding
\textit{OpenMP} was not reflected in the actual implementations and
submissions.  We assert that \textit{OpenMP} is still not as easy as it appears
to be, and that the students' responses strengthen the argument for teaching
lower-level technologies for parallel and distributed computing.
The main contribution of this paper is the conclusion that students
are able to achieve better performance results using \textit{Pthreads}
than using \textit{OpenMP}, despite their perception that \textit{Pthreads}
is harder to learn and use.

Brown \textit{et al.}~\cite{Brown:2010:SPC:1971681.1971689} points that the
choice of a software tool should be only one of the topics explored during a
parallel programming course.  In future work we will compare the perception and
performance of students regarding learning and achieving performance
improvements using different languages and tools for parallel and distributed
computing.  Since we only studied an embarrassing parallel problem which could
give an initial advantage to \textit{OpenMP}, in future work we will explore
how more difficult parallel and distributed problems impacts the students'
perceptions and performance.
